---
title: "Lab 6 - Alternative"
subtitle: "Contributing to R for Data Science"
editor: source
---

The exercises from the functions below were pulled from the newest version of
[*R for Data Science*](https://r4ds.hadley.nz/). Specifically, from Chapters 25 
and 26. For this "alternative" lab you will complete the exercises from the 
textbook, with the option of submitting a pull request to the 
[repository for the textbook solutions](https://github.com/mine-cetinkaya-rundel/r4ds-solutions/).

# Vector Functions

**Question 1:** The `rescale01()` function below performs a min-max scaling to 
standardize a numeric vector, but infinite values are left unchanged. Rewrite
`rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1?
*Hint: This seems like a great place for `case_when()`!

```{r}
#| label: question-1

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
```

**Question 2:** Write a function that accepts a vector of birthdates and 
computes the age of each person in years.

```{r}
#| label: question-2

```

**Question 3:** Write a function that computes the variance and skewness of a
numeric vector. Feel free to look up the definitions on Wikipedia or elsewhere!

```{r}
#| label: question-3

```

**Question 4:** Write a function called `both_na()` which takes two vectors of
the same length and returns the number of positions that have an `NA` in both
vectors.

## Data Frame Functions

**Question 5:** Insert the data frame function you wrote from Lab 6 (either
Exercise 1 or Exercise 2). 

```{r}
#| label: function-from-lab-2

```

For Questions 6 - 10 you will write different functions which work with data 
similar to the `nycflights13` data. 

**Question 6:** Write a `filter_severe()` function that finds all flights that
were cancelled (i.e. `is.na(arr_time)`) or delayed by more than an hour.

```{r}
#| label: question-6

```

**Question 7:** Write a `summarize_severe()` function that counts the number of 
cancelled flights and the number of flights delayed by more than an hour.

```{r}
#| label: question-7

```

**Question 8:** Modify your `filter_severe()` function to allow the user to 
supply the number of hours that should be used to filter the flights that were
cancelled or delayed. 

```{r}
#| label: question-8

```

**Question 9:** Write a `summarize_weather()` function that summarizes the
weather to compute the minimum, mean, and maximum, of a user supplied variable. 

```{r}
#| label: question-9

```

**Question 10:** Write a `standardize_time()` function that converts the user
supplied variable that uses clock time (e.g., `dep_time`, `arr_time`, etc.) into
a decimal time (i.e. hours + (minutes / 60)).

```{r}
#| label: question-10

```

# Plotting Functions

You might want to read over the [Plot Functions section of *R for Data Science*](https://r4ds.hadley.nz/functions.html#plot-functions)

**Question 11:** Build a `sorted_bars()` function which:

- takes a data frame and a variable as inputs and returns a **vertical** bar
chart 
- sorts the bars in decreasing order (largest to smallest)
- adds a title that includes the context of the variable being plotted

*Hint 1: The `fct_infreq()` and `fct_rev()` functions from the forcats package will be helpful for sorting the bars!*
*Hint 2: The `englue()` function from the rlang package will be helpful for adding a variable's name into the plot title!*

```{r}
#| label: question-11

```

# Iteration

Alright, now let's take our plotting function and iterate it! 

**Question 12:** Make a sorted barplot for every character variable in the `mpg`
dataset (built into `ggplot2`). 

```{r}
#| label: question-12

```

# Contributing to the R for Data Science Community!

The functions you wrote for exercises 1-10 came from *R for Data Science*. You
could consider making a pull request to the repository for the solutions! 

<https://github.com/mine-cetinkaya-rundel/r4ds-solutions>

To learn more about how to make a pull request I would suggest this article: <https://usethis.r-lib.org/articles/pr-functions.html>
